Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-15 text/back_end_KS.qmd
chicago results
smooth rq1 results chi

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. Chicago Geomarkets follow a classic geographical structure of an inner city core surrounded by a ring of suburbs, as described by @RN4982 in the development of The Market Segment Model. Northern suburb Geomarkets-Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie- experienced relatively small population growth from 1980 through 2020. The greatest population growth across the metropolitan area from 1980 through 2020 occurred in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago, which encapsulates the center core of the metropolitan area, was the only Geomarket to experience population declines. Despite remaining the most densely populated Geomarket, The City of Chicago population decreased from 3.1 Million in 1980 to 2.8 Million in 2020.

These population changes have resulted in shifting racial and ethnic demographics within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each of the seven Chicago Geomarkets across three census periods. In 1980, all Geomarkets were predominantly White, ranging from 83.1% in the South and Southwest Suburbs to approximately 93.3% in the North Shore—except for the City of Chicago, where the White population made up only 45.5%. By 2020, many Geomarkets that were previously predominantly White shifted to a nearly balanced distribution between White residents and Residents of Color, including Chain of Lakes (58.7% White, 7.6% Black, 23.5% Hispanic; 7.4% Asian Pacific Islander), Northwest Suburbs (61.5% White, 3% Black, 18.2% Hispanic; 14.8% Asian Pacific Islander), Evanston and Skokie (63.7% White, 6.7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (55.6% White, 9.3% Black, 23.4% Hispanic; 9.2% Asian Pacific Islander), and South and Southwest Suburbs (53.4% White, 23.8% Black, 17.9% Hispanic; 2.8% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic demographics was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7.4 percentage point increase for the Evanston and Skokie Geomarket to a 19.2 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie).

@fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. For instance, North Shore experienced some increases in Hispanic (2.2.% in 1980 to 7.9% in 2020), Asian Pacific Islander (6.6% in 2000 to 9.9% in 2020), and Black (1.8.% in 1980 to 2.6% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (36.6% in 1980 to 32.9% in 2020), alongside increases in its Hispanic (14.8% in 1980 to 25.8% in 2020) and Asian Pacific Islander populations (4.0% in 2000 to 6.1% in 2020), further solidifying it as a majority resident of color Geomarket.

@@

@fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (12.5%). However, by 2020, it experienced the largest percentage point increase, reaching 37.6%. In comparison, all other Geomarkets experienced smaller gains in the share of residents with a bachelor’s degree or higher over the same period. These trends contributed to a narrowing of the educational attainment gap across Geomarkets from 1980 to 2020.

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospecs across the entire state of Illinois. Two of these orders applied the same filters: relatively lower SAT score thresholds ranging from 1020 to 1150, GPA ranging from B- to A+, and high school class targeting students from the 2019-2020 and 2020-2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds- two targeted students with middle-range scores (1160 to 1300) and two targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can use prospects within these orders coming from the Chicago metropolitan area and simulate student lists if Geomarket filters were applied across low, mid, and high range score thresholds.

@@

These Geomarket contribution patterns become more pronounced as SAT score thresholds increase. For instance, @fig-chicago-rq2-midSAT-combo presents prospect pools disaggregated by both first-generation college student status and race/ethnicity within the middle-range SAT score (1160–1300) orders. The City of Chicago Geomarket still contributes a disproportionately larger share of first -generation college students - both those whose parents did not attend college and those whose parents attended some college but did not complete a degree - across most racial and ethnic groups. However, the magnitude of this disproportionately for City of Chicago becomes even greater for Asian students, with a nearly 30-percentage-point difference collectively for all first-generation college students (35% no college and 22% some college) relative to their overall Geomarket representation (28%). Similar to patterns observed at lower SAT score thresholds, the Geomarket contributions to first-generation Hispanic college students remain relatively balanced. All Geomarkets, with the exception of North Shore (22%) and Evanston and Skokie (32%), contributed more than 50% first-generation college students across both parental education categories to the overall prospect pool.

>>>>

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect a classic urban geographical structure - an inner-city core surrounded by a ring of suburbs - as described by @RN4982 in the development of The Market Segment Model. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area -Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occured in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.1 Million in 1980 to 2.8 Million in 2020, despite remaining the most densely populated Geomarket.

These population shifts have contributed to changes in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarkets across the three census periods. In 1980, all Geomarkets —except for the City of Chicago - were predominantly White, with proportions ranging from 83.1% in the South and Southwest Suburbs to approximately 93.3% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 45.5% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more balanced racial/ethnic distribution, including Chain of Lakes (58.7% White, 7.6% Black, 23.5% Hispanic; 7.4% Asian Pacific Islander), Northwest Suburbs (61.5% White, 3% Black, 18.2% Hispanic; 14.8% Asian Pacific Islander), Evanston and Skokie (63.7% White, 6.7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (55.6% White, 9.3% Black, 23.4% Hispanic; 9.2% Asian Pacific Islander), and South and Southwest Suburbs (53.4% White, 23.8% Black, 17.9% Hispanic; 2.8% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7.4 percentage point increase for the Evanston and Skokie Geomarket to a 19.2 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie).

@fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. For instance, North Shore experienced some increases in Hispanic (2.2.% in 1980 to 7.9% in 2020), Asian Pacific Islander (6.6% in 2000 to 9.9% in 2020), and Black (1.8.% in 1980 to 2.6% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (36.6% in 1980 to 32.9% in 2020), alongside increases in its Hispanic (14.8% in 1980 to 25.8% in 2020) and Asian Pacific Islander populations (4.0% in 2000 to 6.1% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area.

@@

@fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (12.5%). However, by 2020, the Geomarket experienced the largest percentage point increase in college degree attainment, reaching 37.6%. In comparison, all other Geomarkets experienced smaller gains in the share of residents with a bachelor’s degree or higher over the same period. These trends contributed to a narrowing of the educational attainment gap across Geomarkets from 1980 to 2020.

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospects across the entire state of Illinois. Two of these orders applied the same filters: relatively lower SAT score thresholds ranging from 1020 to 1150, GPA ranging from B- to A+, and high school class targeting students from the 2019-2020 and 2020-2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds- two targeted students with middle-range scores (1160 to 1300) and two targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can utilize prospects from the Chicago metropolitan area within the resulting student lists to examine Geomarket contributions by simulating the application of a Geomarket filter across low, mid, and high range score threshold orders.

@@

Geomarket contribution patterns become more pronounced as SAT score thresholds increase. For instance, @fig-chicago-rq2-midSAT-combo presents prospect pools disaggregated by both first-generation college student status and race/ethnicity within the middle-range SAT score (1160–1300) orders. The City of Chicago Geomarket still contributes a disproportionately larger share of first -generation college students - both those whose parents did not attend college and those whose parents attended some college but did not complete a degree - across most racial and ethnic groups. However, the magnitude of this disproportionately for City of Chicago becomes even greater for Asian students, with a nearly 30-percentage-point difference collectively for all first-generation college students (35% no college and 22% some college) relative to their overall Geomarket representation (28%). Similar to patterns observed at lower SAT score thresholds, the Geomarket contributions to first-generation Hispanic college students remain relatively balanced. All Geomarkets, with the exception of North Shore (22%) and Evanston and Skokie (32%), contributed more than 50% first-generation college students across both parental education categories to the overall prospect pool.
